\pkg{sdmTMB}: An \proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields

### 2023-09-18 doc/paper-jss/sdmTMB-paper.Rnw
SVCs
Edits

<<libs-extras>>=
theme_set(theme_light())
options(ggplot2.continuous.fill = "viridis")
@

@@

We can use the \texttt{tidy()} function (formatted to match the \pkg{broom}
package TODO) to obtain a data frame with parameter estimates (standard methods
such as \texttt{fixef()} and \texttt{vcov()} are also available). The standard
errors on our fixed effects have increased with the spatial random field:

@@

We can do additional model comparison with cross validation via
\texttt{sdmTMB\_cv()}. Here we will do 10-fold cross validation with the folds
constructed randomly. Using the \texttt{fold\_ids} argument, we could supply our
own folds and conduct spatially blocked cross validation. If we set a parallel
plan with the \pkg{future} package \citep{bengtsson2021future}, our folds will
be fit in parallel.

@@

\sloppy To visualize our model, we can make predictions with the
\texttt{predict()} method (\texttt{?predict.sdmTMB}) and optionally use the
\texttt{newdata} argument to predict on a new data frame with any locations and
values for the predictor columns. Here, we will predict on a 2 \(\times\) 2 km
grid (\texttt{qcs\_grid}) that covers the entire region of interest so we can
visualize the predictions spatially. The grid contains spatial covariate columns
and all predictors used in the model set at values for which we want to predict.
The output of \texttt{predict()} is a data frame containing overall estimates in
link space (\texttt{est}), estimates from the non-random-field components
(\texttt{est\_non\_rf}; here, intercept and depth), and estimates from the
individual random field components (here, \texttt{omega\_s}---the spatial
field).

@@

<<pcod-predict-maps, fig.width=10, fig.asp=0.4, out.width="6.1in", fig.cap="Prediction components from the binomial species distribution model of Pacific Cod. Shown are (a) the quadratic effect of bottom depth, (b) the spatial random field in link (logit) space, and (c) the overall prediction, which here is the combination of panels a and b. Note the difference between predictions from depth alone (a) and predictions including a spatial random field (c).">>=
plot_spatial_map <- function(dat, column, title) {
  ggplot(dat, aes(X, Y, fill = {{ column }})) +
    geom_raster() +
    coord_fixed() +
    theme(legend.position= "bottom") +
    ggtitle(title) +
    theme(axis.title = element_blank(), axis.ticks = element_blank(), axis.text = element_blank())
}
g1 <- plot_spatial_map(p, plogis(est_non_rf), "(a) Fixed effects")
g2 <- plot_spatial_map(p, omega_s, "(b) Spatial random field") +  scale_fill_gradient2()
g3 <- plot_spatial_map(p, plogis(est), "(c) Combined prediction")
cowplot::plot_grid(g1, g2, g3, ncol = 3)
@

@@

\[
\begin{aligned}
  \gamma_{t=1} &\sim \operatorname{Normal} \left(0, \sigma^2_{\gamma} \right),\\
  \gamma_{t>1} &\sim \operatorname{Normal} \left(\rho_\gamma\gamma_{t-1}, \sqrt{1 - \rho_\gamma^2} \sigma^2_{\gamma} \right).
\end{aligned}
\]

<<dog-tw, results='hide', message=FALSE, warning=FALSE, echo=TRUE>>=
fit_tw <- sdmTMB(
  catch_weight ~ s(log_depth),
  data = dat,
  mesh = mesh,
  offset = log(dat$area_swept),
  time = "year",
  time_varying = ~ 1,
  time_varying_type = "ar1",
  spatial = "on",
  spatiotemporal = "iid",
  anisotropy = TRUE,
  silent = FALSE,
  family = tweedie()
)
@

@@

We find that the Poisson-link delta-lognormal model is favoured by AIC.

@@

Similarly to the first example, we can visualize model predictions on a grid
covering the area of interest. Because this is a spatiotemporal model, we first
need to replicate our grid for each year we will predict on. Since this is
a common operation, we include the convenience function \texttt{replicate\_df()}
to replicate a data frame. We then ensure our data frame contains all the
predictors used in the model (here \texttt{log\_depth}).

@@

Our prediction data frame is similar to the binomial spatial model, but includes
columns for the two delta model components (labelled with suffixes \texttt{1}
and \texttt{2}) and adds an \texttt{epsilon\_st} column for spatiotemporal
random effects. We can easily generate plots from this data frame using code
similar to in our spatial example with Pacific Cod. We suppress that code for
brevity.

@@

In our final example, we demonstrate a model with spatially varying coefficient
(SVC) effects and illustrate combining uncertainty from parameters by working
with draws from the joint parameter precision matrix. Spatially varying
coefficients \ldots{} TODO REFS. In addition, we illustrate using restricted
maximum likelihood (REML), which integrates over the fixed effects as well as
the random effects.

>>>>

<<libs-extras, cache=FALSE, echo=FALSE>>=
theme_set(theme_light())
options(ggplot2.continuous.fill = "viridis")
@

@@

We can use the \texttt{tidy()} function \citep[formatted to match the
\pkg{broom} package;][]{broom} to obtain a data frame with parameter estimates
(standard methods such as \texttt{fixef()}, \texttt{confint()}, and
\texttt{vcov()} are also available). The standard errors on our fixed effects
have increased with the spatial random field:

@@

We can do additional model comparison with cross validation via
\texttt{sdmTMB\_cv()}. We will do 10-fold cross validation with the folds
constructed randomly. Using the \texttt{fold\_ids} argument, we could supply our
own folds and conduct spatially blocked cross validation. If we set a parallel
plan with the \pkg{future} package \citep{bengtsson2021future}, our folds will
be fit in parallel.

@@

\sloppy To visualize our model, we can make predictions with the
\texttt{predict()} method (\texttt{?predict.sdmTMB}) and optionally use the
\texttt{newdata} argument to predict on a new data frame with any locations and
values for the predictor columns. Here, we will predict on a 2 \(\times\) 2 km
grid (\texttt{qcs\_grid}) that covers the entire region of interest so we can
visualize the predictions spatially. The grid contains spatial covariate columns
and all predictors used in the model set at values for which we want to predict.
The output of \texttt{predict()} is a data frame containing overall estimates in
link space (\texttt{est}), estimates from the non-random-field components
(\texttt{est\_non\_rf}; here, intercept and depth), and estimates from the
individual random field components (here, \texttt{omega\_s}---the spatial
field). We can plot these with \texttt{geom\_raster()} from the \pkg{ggplot2}
package (Figure~\ref{fig:pcod-predict-maps}).

@@

<<pcod-predict-maps, fig.width=10, fig.asp=0.4, out.width="6.1in", fig.cap="Prediction components from the binomial species distribution model of Pacific Cod. Shown are (a) the quadratic effect of bottom depth, (b) the spatial random field in link (logit) space, and (c) the overall prediction, which here is the combination of panels a and b. The spatial random field represents spatially correlated latent effects not accounted for by the fixed effects. Note the difference between predictions from depth alone (a) and predictions including a spatial random field (c).">>=
plot_spatial_map <- function(dat, column, title) {
  ggplot(dat, aes(X, Y, fill = {{ column }})) +
    geom_raster() +
    coord_fixed() +
    theme(legend.position= "bottom") +
    ggtitle(title) +
    theme(axis.title = element_blank(), axis.ticks = element_blank(), axis.text = element_blank())
}
g1 <- plot_spatial_map(p, plogis(est_non_rf), "(a) Fixed effects")
g2 <- plot_spatial_map(p, omega_s, "(b) Spatial random field") +  scale_fill_gradient2()
g3 <- plot_spatial_map(p, plogis(est), "(c) Combined prediction")
cowplot::plot_grid(g1, g2, g3, ncol = 3)
@

@@

\[
\begin{aligned}
  \gamma_{t=1} &\sim \operatorname{Normal} \left(0, \sigma^2_{\gamma} \right),\\
  \gamma_{t>1} &\sim \operatorname{Normal} \left(\rho_\gamma\gamma_{t-1}, \sqrt{1 - \rho_\gamma^2} \sigma^2_{\gamma} \right),
\end{aligned}
\]

where \(\rho_\gamma\) represents the correlation between intercepts \(\gamma\)
at time $t-1$ and $t$.

We can fit this model as:

<<dog-tw, results='hide', message=FALSE, warning=FALSE, echo=TRUE>>=
fit_tw <- sdmTMB(
  catch_weight ~ s(log_depth),
  data = dat,
  mesh = mesh,
  family = tweedie(),
  offset = log(dat$area_swept),
  time = "year",
  time_varying = ~ 1,
  time_varying_type = "ar1",
  spatial = "on",
  spatiotemporal = "iid",
  anisotropy = TRUE,
  silent = FALSE
)
@

@@

We find that the Poisson-link delta-lognormal model is favoured by AIC. In an
applied situation we would consider comparing models with cross validation and
inspecting the distribution of residuals.

@@

Similarly to the first example, we can visualize model predictions on a grid
covering the area of interest. Because this is a spatiotemporal model, we first
need to replicate our grid for each year we will predict on. Since this is
a common operation, we include the function \texttt{replicate\_df()} to
replicate a data frame. We then ensure our data frame contains all the
predictors used in the model (here \texttt{log\_depth}).

@@

Our prediction data frame is similar to the binomial spatial model, but includes
columns for the two delta model components (labelled with suffixes \texttt{1}
and \texttt{2}) and adds an \texttt{epsilon\_st} column for spatiotemporal
random effects. We can easily generate plots from this data frame using
\pkg{ggplot2} code with \texttt{geom\_raster()} similar to our spatial example
with Pacific Cod (Figure~\ref{fig:dog-wcvi-pred}). We suppress that code for
brevity.

@@

In our final example, we demonstrate a model with spatially varying coefficient
(SVC) effects and illustrate combining uncertainty from parameters by working
with draws from the joint parameter precision matrix. Spatially varying
coefficient models \citep{hastie1993} are a class of models in which
coefficients are allowed to vary spatially constrained by some smooth function.
In addition, we illustrate using restricted maximum likelihood (REML), which
integrates over the fixed effects as well as the random effects.
